---
title: Distribution of Urban Trees in Buffalo (Part 1)
date: 2018-07-16
output:
  html_document:
    css: style.css
    highlight: pygments
---

```{r,echo=FALSE}
options(tidyverse.quiet=TRUE)
```

# Introduction

This is part of a series of posts where I examine data sets available as part of the City of Buffalo's [Open Data portal](https://data.buffalony.gov/). Previously I examined recycling data ([post 1](https://kenkellner.com/blog/buffalo-recycle-part1.html) and [post 2](https://kenkellner.com/blog/buffalo-recycle-part2.html)).

The next dataset that caught my eye is a very impressive [tree inventory](https://data.buffalony.gov/Quality-of-Life/Tree-Inventory/n4ni-uuec) from the Bureau of Forestry. It contains information on the species, size, location, and economic contributions of every single street tree in the city.

# Get the Tree Inventory Data

First, I will load the required `R` libraries.
I'll be using `RSocrata` to access the Socrata API in `R`, the `tidyverse` family of packages to manipulate the data, `sf` to handle spatial data, and `ggplot2` for figures.

```{r,warning=FALSE,message=FALSE}
library(RSocrata)
library(tidyverse)
library(sf)
library(ggplot2)
```

Identify the URL for the tree inventory dataset in CSV form:

```{r}
buf_datasets = ls.socrata('https://data.buffalony.gov')
ind = which(buf_datasets$title == 'Tree Inventory')
buf_datasets$distribution[[ind]]$mediaType
csv_url = buf_datasets$distribution[[ind]]$downloadURL[1]
csv_url
```

Then load the dataset into `R` using `RSocrata` as a tibble:

```{r,eval=F}
trees_raw = read.socrata(csv_url) %>% as_tibble()
trees_raw
```

```{r,echo=F}
#Cache the dataset if necessary
if (file.exists('../data/trees_raw.Rdata')){
  load('../data/trees_raw.Rdata')
} else {
  dir.create('../data')
  trees_raw = read.socrata(csv_url) %>% as_tibble()
  save(trees_raw,file='../data/trees_raw.Rdata')
}
trees_raw
```

A quick look at the first 10 rows reveals that there are rows in the dataset for stumps (`STUMP`) and for, presumably, locations where a tree *could* be planted (`VACANT`).
This is potentially interesting information but not for what I'm planning to do with this dataset.
Thus, I removed these rows:

```{r}
trees = trees_raw %>%
  filter(! Common.Name %in% c('STUMP', 'VACANT')) %>%
  rename(Species = Common.Name,
         Eco_Value = Total.Yearly.Eco.Benefits....)
```

# Initial Visualizations

First I wanted to get a rough idea of the spatial distribution of the trees.
My first approach was to convert the `trees` dataset to a simple feature (`sf`) object.
However, I found that `geom_sf()`, the `ggplot2` function normally for plotting `sf` objects, was extremely slow for large spatial point datasets (like this dataset).
Thus, I instead used a combination of `geom_point()` and `coord_map()`, as suggested [here](https://github.com/tidyverse/ggplot2/issues/2718).

```{r}
trees %>%
  ggplot(aes(x = Longitude, y = Latitude, color = Species)) +
  geom_point(size=0.4) +
  coord_map() +
  theme(legend.position='none')
```

It is clear from this figure that the trees are primarily along roads.
Each tree species is represented by a different color, but I had to hide the legend because there are so many species represented in the dataset.

It's easier to visualize distributions of different species if the figure is limited to only the most common species (in this case, trees with more than 1500 individuals):

```{r}
#Make list of most common species
common_species = trees %>%
  group_by(Species) %>%
  summarize(n=n()) %>%
  filter(n>1500) %>%
  pull(Species)

common_species

#Filter dataset and plot
trees %>%
  filter(Species %in% common_species) %>%
  ggplot(aes(x = Longitude, y = Latitude, color = Species)) +
  geom_point(size=0.4) +
  coord_map() +
  guides(colour = guide_legend(override.aes = list(size=2)))
```

Several varieties of maples are among the most common trees in Buffalo.
Another striking pattern here is the concentration of elms (specifically, the Christine Buisman cultivar of *Ulmus minor*) near Soldier's Circle.
Buisman is known for discovering the cause of Dutch elm disease and developing resistant varieties.

# Visualizing Density, Species Richness, and Economic Value

To expand on the basic visualizations above, I wanted to generate visualizations of three variables derived from the tree dataset: (1) tree density; (2) tree species richness (that is, the number of unique species present) per unit area; and (3) total economic value created by urban trees per unit area.
To accomplish this, I divided up the Buffalo region into 600 roughly square-shaped cells of equal area (20x30), and calculated the value of the three variables within each cell - essentially a crude rasterization.

## Creating the (Raster) Cell Boundaries

The first step was to convert the `tree` dataset into an `R` simple feature, so I could manipulate it with the tools in package `sf`.
The dataset included columns with the precise latitude and longitude of each tree, which I then transformed to a UTM projection.

```{r}
trees_proj = trees %>%
  st_as_sf(coords = c("Longitude", "Latitude"),
          crs = "+proj=longlat +datum=WGS84") %>%
  st_transform(crs="+proj=utm +zone=17 +datum=WGS84")
```

Next I used the `sf` function `st_make_grid()` to construct a 20x30 raster covering the Buffalo area (more specifically, the area of the `trees` dataset).
The spatial information of the resulting simple feature was in a column named `geom`, which I renamed `geometry` so it would be recognized properly when creating figures later.
Finally I added two new columns: one that contained a unique ID for each cell (the row number) and one containing the cell area for use in future calculations.

```{r}
trees_grid = st_sf(geom=st_make_grid(trees_proj,n=c(20,30))) %>%
              st_transform(crs="+proj=utm +zone=17 +datum=WGS84") %>%
              rename(geometry = geom) %>%
              rownames_to_column() %>%
              mutate(area = st_area(.))
```

Here's the point density plot of crabapple trees, divided up by the grid cell map:

```{r}
trees_proj %>% 
  filter(Species == 'CRABAPPLE') %>% 
  ggplot() + geom_sf(data=trees_grid) +
  geom_sf()
```

## Calculating Variable Values in Each Cell

With the raster cell boundaries in hand, the next step was to identify which specific cell each tree was located inside, using `st_intersection()`:

```{r,warning=FALSE}
trees_proj = trees_proj %>%
  st_intersection(trees_grid)
```

With each tree assigned to a single grid cell, I calculated the three summary statistics for each grid cell:

```{r}
tree_stats = trees_proj %>%
  st_set_geometry(NULL) %>%
  group_by(rowname) %>% #Unique cell ID
  summarize(richness = length(unique(Species)),
            n = n(), #Count
            density = as.numeric(n/(min(area)/10000)), #Density in trees/ha
            #Economic value in $ per hectare
            eco_val = as.numeric(sum(Eco_Value)/(min(area)/10000))) 

tree_stats
```

Finally, I joined these summary stats to the original `trees_grid` raster to allow them to be used in figures.

```{r}
trees_grid = trees_grid %>%
  left_join(tree_stats, by='rowname')
```

## Plotting the Results

To visualize the spatial distribution of tree density, species richness, and economic value across the city, I built a series of figures in which the color of a grid cell corresponds to the value of the variable of interest.

First, I set a common `ggplot2` theme for each figure to avoid repeating code:

```{r}
plot_theme = theme(axis.ticks.x = element_blank(),
                   axis.text.x = element_blank(),
                   axis.ticks.y = element_blank(),
                   axis.text.y = element_blank())
plot_cols = scale_fill_gradient(low='red', high='green')
```

Next, I generated each raster plot and then combined them into a single figure for easy comparison:

```{r}
#Tree density plot
dens_plot = trees_grid %>%
  ggplot() + 
  geom_sf(aes(fill=density)) +
  labs(fill='Density \n(trees/hectare)') +
  plot_theme + plot_cols

#Tree richness plot
rich_plot = trees_grid %>%
  ggplot() + 
  geom_sf(aes(fill=richness)) +
  labs(fill='Number of tree \nspecies present') +
  plot_theme + plot_cols

#Economic value plot
ecoval_plot = trees_grid %>%
  ggplot() + 
  geom_sf(aes(fill=eco_val)) +
  labs(fill='Tree economic \nvalue ($/hectare)') +
  plot_theme + plot_cols

#Combine into multi-panel plot for comparison
gridExtra::grid.arrange(dens_plot,rich_plot,ecoval_plot,ncol=2)
```

For all three variables, there appear to be two clusters of high values that I guessed where associated with larger city parks.

To test this hypothesis I decided to overlay park boundaries on top of the richness figure.
Luckily Buffalo OpenData has spatial information for parks available in the form of a shapefile (as part of the dataset titled `'Parks'`).
However, the `RSocrata` package does not have a convenient way of reading shapefile data directly into R the way it does for CSVs.
To facilitate this and future similar tasks I wrote a small function `get_shapefile()` that downloads a zipped shapefile from a provided URL, unpacks it into a temporary directory, and reads it into R as a simple feature.

```{r}
get_shapefile <- function(url){
  dest_dir = tempdir()
  dest <- tempfile(tmpdir=dest_dir)
  download.file(url, dest, quiet=T)
  unzip(dest, exdir = dest_dir)
  shape_name = grep('.shp',list.files(dest_dir),value=T)
  setwd(dest_dir)
  sf::st_read(shape_name,quiet=TRUE) 
}
```

I grabbed the shapefile URL using `ls.socrata()` and fed it to the new function:

```{r}
ind = which(buf_datasets$title == 'Parks')
shape_url = buf_datasets$distribution[26][[1]]$downloadURL[3]
park_bounds = get_shapefile(shape_url)
```

Then I overlaid the park boundaries on the species richness raster:

```{r}
rich_plot + 
  geom_sf(data=park_bounds %>% 
          filter(acres > 5),
        alpha=0.6)
```

As predicted, areas of high tree species richness tend to be located inside parks.

# Future Directions

So far I have only looked at broad spatial patterns in Buffalo trees.
Next I plan to more formally examine variables that affect the spatial distribution of trees, using a Poisson point process model.

I've identified several papers that looked at how economic and demographic variables impact urban trees (for example, [this one](http://journals.sagepub.com/doi/abs/10.1068/a41236) from Tampa, FL).
As with my analysis of the Buffalo [recycling data](https://kenkellner.com/blog/buffalo-recycle-part2.html), the American Community Survey (ACS) should be a good source for these variables.
