---
title: Comparing Recycling Rates Among Buffalo Neighborhoods (Part 1)
date: 2018-05-13
output:
  html_document:
    css: style.css
    highlight: pygments
---

#Introduction

I've long been a stubborn proponent of base `R` for processing and visualizing data.
It remains an incredibly robust and flexible tool for these purposes.
However, I've recently taken a deep dive into the [tidyverse](https://tidyverse.org) family of R packages, which perform the same types of tasks in an arguably more streamlined and clear fashion.

My approach to learning the `tidyverse` and the related `ggplot2` visualization package has involved [several](http://r4ds.had.co.nz/) [excellent](https://csgillespie.github.io/efficientR/) [books](http://ggplot2.tidyverse.org/), but I always learn best by actually applying tools to questions I'm interested in.
So, in addition to applications in my [current research](https://github.com/kenkellner), I've been looking for other available datasets to explore.

The city of Buffalo, NY recently created a website called the [Buffalo OpenData](https://data.buffalony.gov/) for publishing data on city services. 
As I'm from the Buffalo region, I decided this was a perfect source for datasets to work with.

#Downloading the Data

The Buffalo OpenData portal uses the Socrata Open Data API (SODA). Luckily, there is an R client for Socrata, `RSocrata`, available on CRAN.

The `RSocrata` package has a variety of functions available, including one for identifying available datasets (`ls.socrata`) and one for downloading them directly into R (`read.socrata`). Here are a few of the available datasets on the Buffalo OpenData Portal:

```{r}
library(RSocrata)
buf_datasets = ls.socrata('https://data.buffalony.gov')
head(buf_datasets$title)
```

I'm going to explore the first dataset listed, which provides monthly recycling rates for different Buffalo neighborhoods. The dataset is available in several formats:

```{r}
ind = which(buf_datasets$title == 'Neighborhood Curbside Recycling Rates')
buf_datasets$distribution[[ind]]$mediaType
```

You can get the URL for the CSV dataset with `$downloadURL`:

```{r}
csv_url = buf_datasets$distribution[[ind]]$downloadURL[1]
csv_url
```

I'll pass the URL to `read.socrata`, then immediately pipe the resulting data frame into a `tibble`. For that (and later data manipulation) I'll also need the `tidyverse` R package.

```{r,message=FALSE,warning=FALSE}
library(tidyverse)
rec_raw = read.socrata(csv_url) %>% as_tibble()
rec_raw
```

Each row of the dataset represents a neighborhood's monthly recycling statistics.
There are columns for neighborhood, date, total recycling, total garbage, and the recycling rate.

Data are available for roughly the past year:

```{r}
min(rec_raw$DATE)
max(rec_raw$DATE)
```

#Tidying the Dataset

## Cleaning up Variables

The first step in cleaning up this dataset is to rename the final three columns, which are a little long.

```{r}
names(rec_raw)[3:5] = c('RECYCLE','GARBAGE','RATE')
```

Two additional things jumped out at me to fix. 
First the neighborhood names were inconsistently capitalized, which could becoming frustrating later on. 
Second, the `RATE` column appeared to be rounded.
To fix these issues I made everything in the `NEIGHBORHOOD` column uppercase, and calculated the recycling rate myself:

```{r}
rec = rec_raw %>%
	mutate(NEIGHBORHOOD = str_to_upper(NEIGHBORHOOD)) %>%
	mutate(RATE = RECYCLE / (RECYCLE + GARBAGE))
```

##Summarizing by Neighborhood

Currently I have monthly recycling stats. For visualization, I decided to summarize information for each neighborhood across all available months.
I also calculated an additional statistic that represented the proportional change in recycling rate for each neighborhood over the time period of the study:

$$
\textrm{Percent Change} = \frac{Rate_{end} - Rate_{start}}{Rate_{start}}
$$

The equivalent function in R:

```{r}
perc_change = function(val,date){
	start = val[which.min(date)]
	end = val[which.max(date)]
	(end - start)/start
}
```

The tools in the `tidyverse` make generating summary statistics for each neighborhood straightforward:

```{r}
rec = rec %>%
	group_by(NEIGHBORHOOD) %>%
	summarize(MnRecycle = mean(RECYCLE,na.rm=T),
			  MnGarbage = mean(GARBAGE,na.rm=T),
			  MnRate = mean(RATE,na.rm=T) * 100,
			  Change = perc_change(RATE, DATE))

rec
```

Now I have one row per neighborhood with summary statistics for each.

#Visualizing the Data

I used the `ggplot2` library for visualizations. Note that the `geom_sf` function I use later to plot spatial data is not available in the current CRAN version of `ggplot2`, so I installed the latest version from Github instead:

```{r,eval=FALSE}
devtools::install_github('tidyverse/ggplot2')
```

##Recycling Rate

Here's a quick and dirty histogram, using `ggplot2`, showing the distribution of recycling rates by neighborhood.
Most neighborhoods have a rate less than 20%.

```{r,warning=FALSE,message=FALSE}
library(ggplot2)
rec %>% ggplot() +
	labs(title = 'Distribution of recycling rates by neighborhood',
		 x = 'Recycling Rate (%)') +
	geom_histogram(aes(x=MnRate),bins=15) +
	geom_vline(xintercept=mean(rec$MnRate),size=2,color='red') +
	geom_text(aes(x=16.5,y=6,
				  label=paste('Overall mean = ',round(mean(rec$MnRate),2),'%',sep='')))
```

##Change in Recycling Rate

Instead looking at proportional change in recycling rate over the time period of the dataset:

```{r,warning=FALSE,message=FALSE}
rec %>% 
	mutate(above0 = ifelse(Change > 0,'Increasing','Decreasing')) %>%
	ggplot() +
	labs(title = 'Change in neighborhood recycling rate, May 2017-April 2018',
		 x = '% Change in Recycling Rate',
		 fill = 'Trend') +
	geom_histogram(aes(x=Change,fill=above0),breaks=seq(-0.15,0.35,0.05)) +
	geom_vline(xintercept=0,size=2,color='white')
```

Most neighborhoods are showing an increasing trend. 

## Seasonal Patterns

Note that the percent change statistic ignores any seasonal patterns in recycling rate that may exist.
Summarizing the raw data by month reveals that recycling rates may be slightly higher in the winter, but the trend is not strong. 

```{r,warning=FALSE,message=FALSE}
library(lubridate) #For handling dates
rec_raw %>%
	mutate(MONTH = factor(month(DATE)),
		   SEASON = ifelse(MONTH%in%c(11,12,1,2), 'Winter','Other')) %>%
	ggplot(aes(x=MONTH,y=RATE, fill=SEASON)) +
		labs(title = 'Recycling rate by month',
			 x = 'Month', y = 'Recycling rate') +
		geom_boxplot()
```

Regardless, this dataset spans almost exactly one year (May 2017 - April 2018) so seasonal effects should be minimized.

#Georeferencing the Neighborhoods

## Finding Spatial Data

To plot these data spatially (and to connect them to other spatially-referenced datasets), I needed information about the shape and location of each neighborhood.
The [metadata](https://data.buffalony.gov/Quality-of-Life/Neighborhood-Curbside-Recycling-Rates/ug79-xatx) for this dataset refers to the neighborhood boundaries in the Buffalo ArcGIS database, but strangely this spatial information does not appear to be available on the OpenData portal.

I was forced to turn to Google instead, and quickly found a database of neighborhood boundary shapefiles provided by [Zillow](https://www.zillow.com/howto/api/neighborhood-boundaries.htm). 
The Zillow shapefiles are at the state level, so I downloaded the zipped-up file for New York and extracted it:

```{r,warning=FALSE,message=FALSE}
src_url = 'https://www.zillowstatic.com/static-neighborhood-boundaries/LATEST/static-neighborhood-boundaries/shp/ZillowNeighborhoods-NY.zip'
if (!file.exists('../data/ZillowNeighborhoods-NY.zip')){
	dir.create('../data')
	download.file(src_url,'../data/ZillowNeighborhoods-NY.zip')
}
unzip('../data/ZillowNeighborhoods-NY.zip',exdir='../data/')
```


To read in the shapefile I used the `sf` package, which can handle multiple different spatial data formats and also supports piping in the `tidyverse` style.
I selected only the neighborhoods in the city of Buffalo, and also converted the neighborhood names to be all uppercase (to match the recycling data).

```{r,warning=FALSE, message=FALSE}
library(sf)
nb = st_read('../data/ZillowNeighborhoods-NY.shp',quiet=T) %>%
	filter(City == 'Buffalo') %>%
	mutate(NEIGHBORHOOD = str_to_upper(Name)) %>%
	select(NEIGHBORHOOD)
```

## Joining the Datasets

The next step was to join the recycling data to the neighborhood spatial data by matching `NEIGHBORHOOD` names. 
I expected that there would be some mismatches to fix. 
I checked which neighborhoods in the spatial dataset didn't have recycling data:

```{r}
nb %>% full_join(rec,by='NEIGHBORHOOD') %>% filter(is.na(MnRecycle)) %>% as_tibble()
```

and which neighborhoods in recycling dataset didn't have spatial data:

```{r}
nb %>% full_join(rec,by='NEIGHBORHOOD') %>% filter(is.na(st_dimension(.))) %>% as_tibble()
```

There are a few obvious issues of inconsistent naming which I fixed before doing a final join of the recycling and spatial datasets.

```{r}
nb = nb %>%
	mutate(NEIGHBORHOOD = recode(NEIGHBORHOOD,
				`MLK PARK` = 'M.L.K. PARK',
				`DELAWARE-WEST FERRY` = 'DELAWARE W. FERRY',
				`BROADWAY-FILLMORE` = 'BROADWAY FILLMORE'))

spatial_rec = nb %>%	
	left_join(rec,by='NEIGHBORHOOD')
```

#Building Maps

Now that the neighborhoods are georeferenced, I can visualize them in space using `ggplot2`.The first step is to calculate centroids for each neighborhood and extract their coordinates.
I need these coordinates later so I can label individual neighborhoods.

```{r,warning=FALSE, message=FALSE}
nb_centers = spatial_rec %>%
	st_centroid() %>%
	mutate(X = st_coordinates(.)[,1],
		   Y = st_coordinates(.)[,2])
```

## Recycling Rate

I used `ggplot` with `geom_sf` to plot the neighborhood polygons and color them according to recycling rate. 
I also added labels to the three highest and three lowest neighborhoods by rate with `geom_label`, using the coordinates I extracted above.

```{r,fig.height=6}
spatial_rec %>%
	ggplot() +
	geom_sf(aes(fill=MnRate)) +
	scale_fill_gradient(low='red',high='green') +
	labs(title = 'Recycling rate in Buffalo neighborhoods',
		 fill = '% Waste \n Recycled',
		 x=NULL, y=NULL) +	
	theme(axis.text.x = element_blank(),
		  axis.ticks.x = element_blank(),
		  axis.text.y = element_blank(),
		  axis.ticks.y = element_blank()) +
	geom_label(data=nb_centers %>% top_n(3,MnRate),
			   mapping=aes(X,Y,label=NEIGHBORHOOD),
			   color='darkgreen',alpha=0.7) +
	geom_label(data=nb_centers %>% top_n(-3,MnRate),
			   mapping=aes(X,Y,label=NEIGHBORHOOD),
			   color='red',alpha=0.7)

```

Lower recycling rates are concentrated in the center of the city, while neighborhoods in the northern part of the city have higher rates.

## Change in Recycling Rate

Buffalo has recently made [a push to improve recycling rates](https://buffalorecycles.org/), so it would also be interesting to see which neighborhoods have shown the greatest improvement in recycling rate over the last year.

I created a nearly identical figure using the percent change in rate statistic I calculated earlier:

```{r,fig.height=6}
spatial_rec %>%
	ggplot() +
	geom_sf(aes(fill=Change)) +
	scale_fill_gradient(low='red',high='green') +
	labs(title = 'Change in recycling rate in Buffalo neighborhoods',
		 fill = '% Change \n in Rate',
		 x=NULL, y=NULL) +	
	theme(axis.text.x = element_blank(),
		  axis.ticks.x = element_blank(),
		  axis.text.y = element_blank(),
		  axis.ticks.y = element_blank()) +
	geom_label(data=nb_centers %>% top_n(3,Change),
			   mapping=aes(X,Y,label=NEIGHBORHOOD),
			   color='darkgreen',alpha=0.7) +
	geom_label(data=nb_centers %>% top_n(-3,Change),
			   mapping=aes(X,Y,label=NEIGHBORHOOD),
			   color='red',alpha=0.7)

```

There's less spatial aggregation in percent change. 
However there seems to be a general trend in which neighborhoods with the greatest increase in recycling rate were among those that had lower overall recycling rates. 
This makes sense as these neighborhoods probably had the most room for quick improvements.

However, a cursory analysis reveals that this trend is weak at best:

```{r,warning=FALSE,message=FALSE}
spatial_rec %>%
	ggplot(aes(x=MnRate,y=Change)) +
	labs(title= 'Mean recycling rate and % change in recycling have a weak relationship',
		   x='Mean Recycling Rate',
		   y='% Change in Recycling Rate') +
	geom_point() +
	geom_smooth(method=lm)
```

#Next Steps

So far I haven't done much with this dataset other than look at surface patterns.
Now that I have the recycling data georeferenced, I plan to model recycling rate as a function of other variables that are spatially referenced - for example, U.S. Census block data.

Read part 2 of this analysis [here](buffalo-recycle-part2.html).
