---
title: 'Recycling in Buffalo Part 2: Effects of Income and Education'
date: 2018-05-21
output:
  html_document:
    theme: journal
    css: style.css
    highlight: pygments
---

```{r,echo=FALSE,warning=FALSE,message=FALSE}
#Data from last post
tempR <- tempfile(fileext = ".R")
library(knitr)
purl("buffalo-recycle-part1.Rmd", output=tempR)
source(tempR)
unlink(tempR)
```

# Introduction

In [part 1](https://kenkellner.com/blog/buffalo-recycle-part1.html) of my analysis of the [Buffalo recycling data]() in `R`, I downloaded the raw dataset, looked at some general patterns, and added geospatial information. 
With this dataset in hand, my next step was to identify variables might be good predictors of a neighborhood's recycling rate.

I was primarily interested in how two variables might affect neighborhood recycling rate:

1. Income, and
2. Education level

Unfortunately, neither the recycling data itself nor the neighborhood boundary dataset contain any socioeconomic information about the neighborhoods, so I needed to introduce another dataset. This type of information was not readily available on the Buffalo OpenData portal. Thus, I explored an alternative data source: the U.S. Census Bureau.

# Census Bureau Data

Anyone can download Census data freely from the [website]().
However, this is a rather slow and clunky method.
Instead I obtained a key for API access to the Census data from [here](), and used the `tidycensus` package in R to download census data of interest.

```{r,message=FALSE,warning=FALSE}
library(tidycensus)
```

After obtaining the key you can save it into your `.Renviron` file for later use:

```{r,eval=FALSE}
census_api_key('YOUR KEY HERE', install=TRUE)
```

The package gives you access to two major sources of data collected by the U.S. Census Bureau: the Decennial Census, and the American Community Survey (ACS). 
The Decennial Census data provides reliable information at a very fine scale, but has two primary disadvantages.
First, the last census was in 2010, so the results are nearly 8 years out of date (a big problem in a rapidly changing city like Buffalo).
Second, accessible census information is limited to race, age, and housing data.

On the other hand, the ACS was completed relatively recently (2016), and includes a much wider variety of information including education and income.
The disadvantage is that, being a survey, it does not exhaustively sample the population.
Variables obtained from the ACS will therefore always be *estimates* and have some margin of error.
The finer the spatial scale you drill down to, the greater the margin of error - and I needed a pretty fine spatial scale (neighborhood).
In the end, I decided the advantages of the ACS dataset outweighed this limitation.

# Manipulating the ACS Data

There are thousands of variables in the complete ACS dataset.
You can get a description for all of them using `load_variables`:

```{r,eval=FALSE}
View(load_variables(2016, "acs5", cache = TRUE))
```

After some tedious manual searching, I identified a variable for median income (`B07011_001`)and a set of variables for the number of people 25+ with a given education level (`B15003_*`).
I made a list of these variable codes and gave them slightly better names:

```{r}
get_vars = c('B07011_001',paste('B15003_', sprintf('%03d',c(1,17:25)),sep=''))
names(get_vars) =  c('income','all','hs','ged','some0','some1',
					 'as','bs','ms','prof','phd')
```

When querying the ACS data for these variables, you also need to provide a spatial resolution (e.g., state, county). 
The finest-scale spatial resolution for which I was able to get data was the census tract.
The `get_acs` function from `tidycensus` can be used to extract a set of variables at a given spatial resolution.
I limited the output dataset to Erie County to keep the download small - I'll cut it down to only the Buffalo area later.

```{r,message=FALSE,warning=FALSE}
acs_data = get_acs(geography = 'tract', variables = get_vars,
				   state='NY',county='Erie',geometry='TRUE')

#Look at dataset
as_tibble(acs_data) %>% select(-geometry)
```

As promised, the result is a tidy dataset where each row represents one census tract by variable combination.
The estimate column provides the value for a given variable, and the `moe` column provides the margin of error.
As you can see, there are some pretty large margins of error around some of the variable estimates, no doubt a product of small sample size.
In a serious analysis, the variability in these estimates ought to be accounted for; for the purposes of this post, I will ignore it.

To make the rest of the analysis easier, I needed to split each ACS variable into its own column.
This is very easy with the `spread` function.
I also calculated two summary variables for education: the percent of people in the tract with at least a high school degree, (`atleast_hs`) and the percent of people with at least a Bachelor's degree (`atleast_bs`).

```{r}
library(tidyverse)					   
acs_spread = acs_data %>% 
	select(-moe) %>% 
	spread(key=variable,value=estimate) %>%
	mutate('atleast_hs' = (hs + some0 + some1 + as + bs + ms + prof + phd) / all,
		   'atleast_bs' = (bs + ms + prof + phd) / all) %>%
	select(GEOID,income,atleast_hs,atleast_bs)
```

# Joining the ACS and Recycling Data

The census tracts are roughly equal in size to Buffalo neighborhoods, but the boundaries don't exactly match:

```{r,message=FALSE,warning=FALSE}
library(sf)
library(ggplot2)

thm = theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.text.y = element_blank(), axis.ticks.y = element_blank())

#Dataset from the last post
#Neighborhood boundaries
nb = spatial_rec %>% ggplot() +
	labs(title = 'Neighborhoods') + thm +
	geom_sf()

#Census tract boundaries
tr = acs_data %>%
	st_intersection(spatial_rec) %>%
	group_by(NAME) %>%
	summarize() %>%
	ggplot() +
	labs(title = 'Census Tracts') + thm +
	geom_sf()

gridExtra::grid.arrange(nb,tr, ncol=2)
```

Thus, the ACS data cannot be directly matched to a given neighborhood.
I needed to calculate an average value for each ACS variable across all the tracts that fell in a given neighborhood, weighted by how much of the area of the neighborhood a given tract made up.
My approach to solving this problem was as follows:

1. Calculate the area of each neighborhood using the dataset from the last post (`spatial_rec`).

2. Find out which census tracts overlapped each neighborhood, and what percentage of the area of the neighborhood each of these tracts accounted for (`wt`).

```{r,warning=FALSE,message=FALSE}

step1_2 = spatial_rec %>%
	mutate(tot_area = st_area(.) %>% as.numeric()) %>%
	st_intersection(acs_spread) %>%
	mutate(area = st_area(.) %>% as.numeric(),
		   wt = area / tot_area)
```

3. Use `wt` to get a weighted average of the three ACS variables for each neighborhood.

```{r,warning=FALSE,message=FALSE}
step3 = step1_2 %>%
	group_by(NEIGHBORHOOD) %>%
	summarize(inc = sum(income * wt),
			  hs = sum(atleast_hs * wt * 100),
			  bs = sum(atleast_bs * wt * 100)) %>%
	st_set_geometry(NULL)
```

4. Get the final dataset by joining this data back to the original neighborhood dataset:

```{r,warning=FALSE,message=FALSE}
final = spatial_rec %>%
	left_join(step3, by='NEIGHBORHOOD')

as_tibble(final)
```

# Visualize the Data

The following plot shows the spatial distribution of the income and education variables, along with recycling rate.

```{r}
inc.fig = final %>% ggplot() + thm +
	labs(fill = 'Median\nIncome ($)') +
	scale_fill_gradient(low='red',high='green') +
	geom_sf(aes(fill=inc))
hs.fig = final %>% ggplot() + thm +
	labs(fill = 'Adults With\nHS diploma (%)') +
	scale_fill_gradient(low='red',high='green') +
	geom_sf(aes(fill=hs))
bs.fig = final %>% ggplot() + thm +
	labs(fill = 'Adults with\nBachelors (%)') +
	scale_fill_gradient(low='red',high='green') +
	geom_sf(aes(fill=bs))
rc.fig = final %>% ggplot() + thm +
	labs(fill = 'Recycling\nRate (%)') +
	scale_fill_gradient(low='red',high='green') +
	geom_sf(aes(fill=MnRate))
gridExtra::grid.arrange(inc.fig, hs.fig, bs.fig, rc.fig, ncol=2,
						top='Income, Education, and Recycling Rate are Related')
```

Based on the figure, there appears to be a positive relationship education level and median income (unsurprising). 
Furthermore, more educated and higher-income neighborhoods appear to have a higher recycling rate.
A more formal analysis is needed to confirm these relationships.

# Look for Relationships

Now that I have income and education data for each neighborhood, I can proceed with the actual analysis.


```{r}
final %>%
	ggplot(aes(x = hs, y = MnRate)) +
	geom_point() +
	geom_smooth(method=lm)

test3 %>%
	ggplot(aes(x = bs, y = MnRate)) +
	geom_point()

test3 %>%
	ggplot(aes(x = inc, y = MnRate)) + geom_point()


fit = lm(MnRate ~ inc, data=test3)

plot(test3 %>% filter(!is.na(MnRate)) %>% pull(bs),residuals(fit))

full = lm(MnRate ~ inc + bs, data=test3)
```


