---
title: Fitting Bayesian Models with Integrals
date: 2018-06-03
output:
  html_document:
    css: style.css
    highlight: pygments
---

Recently I've been working on a model that includes a component of the basic form

$$ \sum_{i=1}^n F(x_i)\int_{X}F(x)dx $$

where $F(x)$ is an arbitrary function that includes parameters I want to obtain estimates of.

# Using R

In `R` a maximum likelihood approach to estimation is possible, using the built-in `integrate()` and `optim()` functions:

```{r,eval=FALSE}
#Psuedocode example
#Arbitrary function
F <- function(x, params){...}

loglik <- function(params, data){
	#Integral part
	int_val <- integrate(F, Xmin, Xmax, params)$value
	#Summation part
	sum_val <- 0
	for (i in 1:nrow(data)){		
		sum_val <- sum_val + F(data$x[i], params)
	}
	return(-1 * log(sum(sum_val * int_val)))
}

fit <- optim(inits, loglik, data)
```

This approach works fine.

# Using Popular MCMC Software

In my specific case, however, I would prefer to fit the model in a Bayesian framework using MCMC.
The usual approaches are limited: `WinBUGS`, `JAGS`, and `Stan` all lack built-in integral functions.
Depending on the complexity of the integral, it may be possible to solve it first and then incorporate it into a model in one of these languages.
Alternatively, you could use some kind of numerical approximation (e.g. a Riemann sum), but that could greatly increase MCMC run time.

The `OpenBUGS` flavor of the `BUGS` language actually does have a `integral` function.
For example:

```{r, eval=FALSE}
#OpenBUGS pseudocode
model {
	C <- 10000 #Large constant		
	F(x) <- ... #Some expression of parameters
	inv_val <- integral(F(x), Xmin, Xmax, tol)

	for (i in 1:nobs){
		sum_val[i] <- ... #Expression from above
		loglik[i] <- -1 * log(sum_val[i] * int_val)
		#Zeroes trick for arbitrary likelihood in BUGS
		mn[i] <- loglik + C
		dummy[i] <- 0
		dummy[i] ~ dpois(mn[i])
	}
	#Priors, etc.
}
```

This works reasonably well in my experience, but has the key limitation that the function $F(x)$ to be integrated has to be able to be expressed in the limited BUGS syntax.
If it's a complex function that might be impossible, as it was in my case.
Furthermore, OpenBUGS can be quite frustrating work with due to its infamously cryptic error messages.

# Using NIMBLE

[NIMBLE](https://r-nimble.org/) is a recent `R` package for hierarchical modeling and MCMC.
It has many features, but the one that caught my attention was its ability to extend existing `BUGS` syntax with user-defined functions and distributions written in `R`.
For me this was a perfect solution: I could fit my model in a Bayesian framework, leverage my existing knowledge of `BUGS`, and also incorporate my own functions in the more flexible `R` language (including functions with integrals!).

For example, suppose I have an arbitrary function in `R`, `my_func()`, which calculates the value of the integral I'm interested in:

```{r,eval=FALSE}
#Pseudocode
my_func <- function(params,Xmin,Xmax){
	F <- function(x){...} #Expression of params
	return(integrate(F,Xmin,Xmax)$value)
}
```

Since it's written in `R`, I have lots of freedom for defining `F(x)`. 
Next I need to make `NIMBLE` aware of this function, and provide additional detail about the inputs and outputs.
The `R` function is wrapped in a call to `nimbleRcall` with a similar name.

```{r,eval=FALSE}
#Pseudocode
Rmy_func <- nimbleRcall(function(params=double(1), Xmin=integer(0), 
						Xmax=integer(0)){}, Rfun='my_func', 
						returnType = double(0), envir=.GlobalEnv)
```

The `BUGS` model (nearly identical to the `OpenBUGS` version above) is then written in a wrapper for `NIMBLE`:

```{r, eval=FALSE}
#Pseudocode
mod <- nimbleCode({
	C <- 10000 #Large constant		
	#Call to my custom function
	int_val <- Rmy_func(params, Xmin, Xmax)

	for (i in 1:nobs){
		sum_val[i] <- ... #Expression from above
		loglik[i] <- -1 * log(sum_val[i] * int_val)
		#Zeroes trick for arbitrary likelihood in BUGS
		mn[i] <- loglik + C
		dummy[i] <- 0
		dummy[i] ~ dpois(mn[i])
	}
	#Priors, etc.
})
```

The only difference is the call to my custom function which calculates the integral, `Rmy_func`, instead of the built-in `integral` function. 
Finally, run `NIMBLE` after providing other necessary inputs (data, initial values, MCMC run info, etc.):

```{r,eval=FALSE}
#Pseudocode
mcmc.out = nimbleMCMC(code = mod, constants = inp_constants,
					  data = inp_data, inits = inp_inits(),
                      monitors = param_names,
					  nchains = 3, niter=3000,nburnin=2500,
                      thin=5,summary =TRUE)
```

Calling the `R` function from `NIMBLE` unsurprisingly has a MCMC speed penalty, but I found run times to be comparable to `OpenBUGS` for equivalent models.
