---
title: Simple Static Site Generator with R, Python, and GNU Make
date: 2018-05-15
output:
  html_document:
    css: style.css
    highlight: pygments
---

# Motivation

I'm hoping to do more blogging in the future, and needed a place to post things.
Over the years I've used a few different blogging platforms, including LiveJournal, Blogger, and most recently, [Jekyll](https://jekyllrb.com/).
I really enjoyed the Jekyll approach of writing plain text and compiling it to a series of fixed (i.e., static) html pages.
There are a number of these so-called static-site generators out there.
Since I mainly work with `R`, [blogdown](https://github.com/rstudio/blogdown) is a good choice and no doubt would have worked just fine.
However, I decided I wanted to attempt to build a basic framework myself.

# Features to Include

1. Version control.

2. Write posts in plain text, ideally with good support for inline code, and then generate decent-looking HTML.

3. An organized, efficient approach to the generation process: keep track of page dependencies and re-build a page to HTML only when a dependency has changed.

5. Auto-generation of an index/table of contents page.

5. Auto-generation of an RSS feed.
   
# Version Control

The whole site is in a `git` repository (you can see it [here](https://github.com/kenkellner/blog)).
Only plain text files are tracked - generated files like HTML, XML are not.
The site is organized into two primary folders: folder `src` contains raw text input files, and folder `build` contains the corresponding output HTML files that make up the actual viewable website.

# Writing Posts

For me (an `R` user), [R markdown](https://rmarkdown.rstudio.com/) was the obvious choice for converting plain text to HTML.
Using a simple syntax you can mix text and nicely-formatted code blocks, run `R` code in the background, and generate visualizations that appear in the HTML file.
It's an amazing tool for creating reproducible analyses.
You are also able to run Python code in `Rmarkdown` using the `reticulate` package (something I hope to do more of the future).

I made a simple R script `build_Rmd.R` to convert `Rmd` files to `html` with the `rmarkdown` package:

```{r,eval=FALSE}
args = commandArgs(trailingOnly = TRUE)
rmarkdown::render(args[1],output_file=paste('../',args[2],sep=''))
```

The `commandArgs` function allows this script to take information from command line standard input (in this case the input and output file names).
This will be needed later.

I kept all the detailed `Rmarkdown` options in the yaml header of each `Rmd` file, which looks basically like this:

```{r,eval=FALSE}
title: A title
date: 2018-01-01
output:
  html_document:
    theme: journal
    css: style.css
    highlight: pygments
```

The header allows me to specify a basic HTML theme (`journal`) which I modified with a custom CSS stylesheet (`style.css`).
The `src` directory also contains a file `_navbar.yml`, which contains links that are put into a simple navigation bar inserted at the top of each page when I compile it with `Rmarkdown`.

# Building the Site

I wanted to explicitly organize the dependency structure of my site. 
For example, building `example-post.html` depends on `example-post.Rmd`, `_navbar.yml` and `style.css`.
Also, Rmarkdown can sometimes take a little while to compile `Rmd` to `html` especially if the document is complicated.
Thus, I wanted to avoid constantly re-compiling all pages and only re-compile a page when its dependencies have changed.

This is a common problem in programming, and a time-tested solution is [GNU Make](https://www.gnu.org/software/make/).
`Make` uses a kind of recipe, a `makefile`, to construct output files from a set of dependencies based on rules you define.
If the dependencies of a particular output file change (e.g., you edit one) and you run `make` again, the output will be re-created.
On the other hand, if the dependencies are unchanged and the output already exists, `make` won't waste time generating it again.

In the first part of the `makefile` for my static site, I set up variables `$SRC` and `$BUILD` corresponding to folders containing text files (`src`) and output html (`build`) respectively.

```{sh, eval=FALSE}
SRC=src
BUILD=build
```

Next the makefile makes a list of the names of all `Rmd` files in the `src` directory, saving it into variable `$RMD_IN`. 
By changing the file extension on each of these `Rmd` files to `html`, and changing their directory from `src` to `build`, the makefile now also has a corresponding list of all the output HTML files it needs to generate, saved in `$RMD_OUT`.

```{sh, eval=FALSE}
RMD_IN = $(wildcard $(SRC)/*.Rmd)
RMD_OUT := $(patsubst $(SRC)/%.Rmd,$(BUILD)/%.html,$(RMD_IN))
```

Next I set up a recipe in the `makefile` to tell `make` how to generate a given `html` file (represented with the wildcard notation `%`) from its corresponding dependencies - the matching `Rmd` file, the navbar file, and the CSS file:

```{sh,eval=FALSE}
$(BUILD)/%.html: $(SRC)/%.Rmd $(SRC)/_navbar.yml $(SRC)/style.css 
	Rscript build_Rmd.R $< $@
```

If either (1) the desired HTML file doesn't exist; or (2) one of the dependencies (specified after the colon `:`) have changed since `make` was last run, then `make` will execute the second line, which runs the `R` script I described earlier for running `Rmarkdown`. 
Otherwise, nothing will happen - the HTML file is already exists and is up-to-date.
No need to build it again!

The code above generates a *given* HTML file. 
I want to build all the HTML files in `$RMD_OUT` when I run `make`, or at least all the HTML files don't exist yet or have updated dependencies.
For that I added another rule called `all`:

```{sh,eval=FALSE}
all: $(RMD_OUT)
	@echo "Done"
```

This is called a "phony" `make` rule because it doesn't explicitly generate a new file (i.e., there's no file on the left-hand side of the colon).
However the rule does have dependencies - the list of all required output HTML files (`$RMD_OUT`).
Thus, running `make all` will force all HTML files to be generated or updated (if necessary) according to the rule above.

# Generating an Index Page

I needed a simple landing page for the blog (`index.html`; see it [here](https://kenkellner.com/blog)) that would list all the posts I've written so far.
I wanted this page to contain two pieces of information: the title of each post (with a link to the page), and the date it was posted, in descending order.
To keep things simple and consistent, I decided to use `R` and `Rmarkdown` for this task as well.

The basic steps are as follows:

1. Search the `src` directory for all `Rmd` files:

    ```{r, eval=FALSE}
    f = list.files(pattern='\\.?md')
    f = f[f!='index.Rmd'] #Exclude this file
    ```

2. Extract the post date and title from each file:

    ```{r,eval=FALSE}
    library(stringr)
    
    get_date = function(filepath){
    	ln = grep('date:',readLines(filepath),value=TRUE)
    	dt = strsplit(ln, ': ')[[1]][2]
    	dt
    }
    
    get_title = function(filepath){
    	ln = grep('title:',readLines(filepath),value=TRUE)
    	t = str_split(ln, ': ', 2)[[1]][2]
    	t = gsub("^'","",t)
    	t = gsub("'$","",t)
    	t
    }

    dates = sapply(f,FUN=get_date)
    titles = sapply(f,FUN=get_title)
    ```

3. Generate links to the HTML pages in markdown format:

    ```{r,eval=FALSE}
    filenames = sub('.Rmd','.html',f) 
    links = paste('[',titles,'](',filenames,')',sep='')
    ```

4. Format and print the dates and linked titles in a neat HTML table:

    ```{r, eval=FALSE}
    tab = data.frame(Date=dates,Title=links,row.names=NULL)
    tab = tab[order(tab$Date,decreasing=T),]
    knitr::kable(tab,row.names=FALSE)
    ```

I decided it would be easiest to generate the index file every time I built the site, regardless if there were any new posts to add. 
Therefore, I added a rule to build it to the `all` recipe in the `makefile`:

```{sh,eval=FALSE}
all: $(RMD_OUT)
	@Rscript build_Rmd.R $(SRC)/index.Rmd $(BUILD)/index.html
	@echo "Done"
```

The index page is built in the same way (using `build_Rmd.R`) the actual blog posts are, and thus has the same navigation bar, styling, etc.

# Generating an RSS feed

I first thought I'd try to create an RSS feed for my blog in `R`, for consistency. 
However, though I found several packages to *read* RSS files, I couldn't find one to *create* an RSS file.

My next choice was a Python module called [python-feedgen](https://github.com/lkiesow/python-feedgen), which I've used in the past. Much like with building the `index.html` page, the approach here was to gather dates, titles, and links for each post and feed them into the module.
I added a second script, `build_rss.py`, to do this, containing the steps below.

1. Load required modules:

    ```{python, eval=FALSE}
    from feedgen.feed import FeedGenerator
    import os
    from datetime import datetime
    ```

2. Find all current posts (i.e., all HTML files in the build directory excluding the index):

    ```{python, eval=FALSE}
    links = os.listdir('build')
    links.remove('index.html')
    ```

3. Create sorted lists of dates, links, and titles for each post. Dates and titles are extracted from the `Rmd` source file.

    ```{python, eval=FALSE}
    src = []
    for i in links:
        src.append('src/'+i.replace('.html','.Rmd'))
    
    dates = []
    for i in src:
        for line in open(i):
            if 'date: ' in line:
                dates.append(line.split(': ')[1].strip('\n'))
    
    titles = []
    for i in src: 
        for line in open(i):
            if 'title: ' in line:
                titles.append(line.split(': ',1)[1].strip('\n'))
    
    dates, links, titles = zip(*sorted(zip(dates, links, titles)))
    ```

4. Initialize the RSS feed object (class `FeedGenerator`) and add metadata:

    ```{python, eval=FALSE}
    fg = FeedGenerator()
    fg.id(leader)
    fg.link(href=leader+'feed.xml', rel='self')
    fg.title('Ken Kellner\'s Blog')
    fg.subtitle(' ')
    fg.language('en')
    ```

5. For each post, add an RSS entry to the object using the dates, links, and titles obtained above:

    ```{python, eval=FALSE}
    leader='https://kenkellner.com/blog/' 

    for i in range(len(dates)):
        fe = fg.add_entry()
        fe.title(titles[i])
        fe.id(leader+links[i])
        fe.link(href=leader+links[i])
        fe.author({'name': 'Ken Kellner', 'email': 'contact@kenkellner.com'})
        fe.description('')
        date_raw = dates[i]+' -0500'
        fe.published(datetime.strptime(date_raw, '%Y-%m-%d %z'))
    ```

6. Output the RSS feed as an XML file (`feed.xml`):

    ```{python, eval=FALSE}
    fg.rss_file('build/feed.xml')
    ```

I added another rule to the `makefile` to generate the RSS feed each time I built the site:

```{sh, eval=FALSE}
all: $(RMD_OUT)
	@Rscript build_Rmd.R $(SRC)/index.Rmd $(BUILD)/index.html > /dev/null 2>&1 
	@python3 build_rss.py
	@echo "Done"
```

# Deploying the Site

I host my websites on a Digital Ocean droplet to which I have full filesystem access.
The fastest way to deploy the site is to simply `rsync` the entire local `build` directory to the `blog` directory of the site on the droplet.
I added another "phony" recipe to the `makefile` to do this:

```{sh, eval=FALSE}
deploy:
	@rsync -r --progress --delete --update build/ \
		kllnr.net:/var/www/kenkellner.com/blog/
```

So, each time I complete I post I run `make all` and then `make deploy`, and my blog is up-to-date!
